---
title: Lectures-Conceptual-Models
author: JochemTolsma
date: '2021-01-18'
slug: lectures
categories:
  - R
  - Conceptual Models
tags: []
#linktitle: "Overview Conceptual Models"

summary: "conceptual model, correlation, explanation, tutorial, R, Lavaan"
lastmod: '`r format(Sys.Date(), "%d-%m-%Y")`'

type: book
weight: 4

 
output:
  blogdown::html_page:
    highlight: "haddock"
    number_sections: no
    self_contained: no
    toc: true
    fig_width: 6
    dev: "svg"

---


<!--set global settings--> 
```{r, globalsettings, echo=FALSE, warning=FALSE}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=100),tidy=TRUE, warning = FALSE, message = FALSE, cache=TRUE, attr.source = ".numberLines", class.source="highlightt", results= "hold")
options(width = 100)
```


<!--copy to clipboard
```{r klippy, echo=FALSE, include=TRUE}
require(klippy)
klippy::klippy(position = c('top', 'left'))
#klippy::klippy(color = 'darkred')
#klippy::klippy(tooltip_message = 'Click to copy', tooltip_success = 'Done')
```
<!---
https://www.w3schools.com/w3css/w3css_buttons.asp
https://www.freecodecamp.org/news/a-quick-guide-to-styling-buttons-using-css-f64d4f96337f/
--->

<!---
<style>
.btn-text-right{
	text-align: right;
}
</style>

<div class="btn-text-right">
<button onclick="window.location.href='static/index.Rmd';">download .Rmd </button>  
</div>

--->

## The Basics  


<iframe src="https://player.vimeo.com/video/501760425" width="640" height="360" frameborder="0" allow="autoplay; fullscreen; picture-in-picture" allowfullscreen></iframe>

{{% alert note %}}
After having watched this video you will know:  

- The common assumptions researchers make when drawing conceptual models
  - chronological order  
  - causality  
  - causal importance    
- That relationships between concepts have directions/valence and that it is good practice to make these explicit.  
- That choosing the right labels for your concepts is important:
  - because only with good labels does it make sense to assign valence to the relations;  
  - labels should be close to both the concept as intended as the concept as measured;  
  - researchers operate in (politically and scientifically) sensitive environments and labels should be (cross-culturally, cross-disciplinary) appropriate.    
{{% /alert %}}

---  


